Multi-Domain Adaptation in Neural Machine Translation with Dynamic Sampling Strategies

### 2022-03-01 2022/eamt/eamt-latex/eamt22.tex

In this paper, we contribute to this line of research in several ways. 
\begin{itemize}
	\item First, we propose a novel framework (\emph{Multi-Domain Automated Curriculum}, MDAC for short), which is a variant of Differentiable Data Selection (DDS) of \newcite{Wang20balancing}, that simultaneously accounts for the domain adaptation and the multi-domain adaptation problems.
	\item We demonstrate MDAC's equivalent performance compared to fine-tuning for domain adaptation; and better performance compared to classical choices of static data sampling strategies.
	\item We show that our variant MDAC mitigate some failures of DDS in handling multi-domain training.
	\item We demonstrate various exotic use cases of differentiable data selection frameworks (MDAC and original DDS), including, domain adaptation using unsupervised clustering instead of using domain label a priori which is barely available in real context; domain adaptation using out-of-domain training data and small in-domain validation data; and bi-domain adaptation in which we are only interested in two domain of the training mixture.
\end{itemize}

Based on these experimental results, our main conclusions are that (a) using MDAC often yields overall performance that is as good as the standard fine-tuning strategy for domain adaptation; (b) MDAC can effectively handle a variety of test situations, from targeting one single domain to the full multi-domain scenario.

@@

\subsection{Multi-domain adaptation}\label{ssec:mda}

We now turn to a more realistic scenario and consider multi-domain adaptation, which aims to train one single system with optimal performance averaged over 6 domains. This setting targets a uniform test distribution $\vlambda^t = \vlambda_0$. In this situation, CL \cite{Zhang19curriculum} does not apply. We therefore only contrast the performance of MDAC, DDS and several fixed training data distribution $\vlambda^l \in \big[ \vlambda_0, \vlambda_{0.25}, \vlambda_{0.5}, \vlambda_{0.75}, \vlambda_{1.0}\big]$, where $\vlambda_{\alpha}$ is defined according to equation~\eqref{mixture:trn}.

@@

Multi-domain adaptation (MDA) aims to develop systems that simultaneously bode well for several domains. Like for DA, techniques for supervised MDA combine one or several ingredients: (a) the specialization of data representations \cite{Kobus17domaincontrol} or of sub-networks \cite{Pham19generic} to differentiate the processing of each domain; (b) the use of adversarial techniques to neutralize differences between domains \cite{Britz17mixing,Zeng18multidomain}; (c) the use of automatic domain identification e.g.\ \newcite{Jiang19multidomain}. Unsupervised MDA is studied in \newcite{Farajian17multidomain}, as an instance of unsupervised DA.

@@

\newcite{Kumar19reinforcement} use reinforcement techniques (deep Q-learning) to learn the curriculum strategy: in this work, complexity corresponds to difficulty levels which are binned using contrastive data selection. The reward is based on the increase of the development set loss that results from the actual data selection strategy.\fyDone{Alert: what do we do during warm-up ?} The same technique is recently applied to multilingual NMT in \cite{Kumar21learning}. \newcite{Zhou20uncertainty} propose another curriculum-based approach which instead relies on \emph{instance uncertainty} as a measure of their difficulty and presents the data sample starting with the easiest (more predictable) first. Another contribution of this paper is an alternative stopping criterium. Closest to our problems, \newcite{Wang20learning-multi} adapt CL for multi-domain adaptation, where an optimal instance weighting scheme is found using Bayesian optimization techniques. Each step consists of (a) weighting instances based on relevance features, (b) fine-tuning a pretrained model using the weighted training set, and is applied iteratively to train a sequence of models. The one that maximizes the development set performance is finally retained.

\section{Conclusion and outlook \label{sec:conclusion}}
\fyDone{Write conclusion}
In this study, we have presented a generic framework to perform a variety of adaptation tasks for machine translation, ranging from conventional supervised domain adaptation to multi-domain adaptation and unseen domain adaptation. By experimenting with all these settings, we have shown that the same algorithm, aimed at automatically finding an effective data sampling scheme during the course of training, could be used in all these situations. This algorithm, we believe, provides us with a more sound approach to (multi-domain) DA than existing heuristics and dispenses with the costly search of optimal meta-parameters. Another contribution of our work is an experimental comparison of recent approaches to dynamic data selection.

>>>>

In this paper, we contribute to this line of research in several ways. 
\begin{itemize}
	\item First, we propose a novel framework (\emph{Multi-Domain Automated Curriculum}, MDAC for short), a variant of Differentiable Data Selection (DDS) of \newcite{Wang20balancing} initially applied to multilingual NMT, that simultaneously accounts for the domain adaptation and the multi-domain adaptation problems.
	\item We demonstrate MDAC's equivalent performance compared to fine-tuning for domain adaptation \ref{ssec:da} and better performance compared to classical choices of static data sampling strategies for multi-domain settings \ref{ssec:mda}.
	\item We show that our variant MDAC mitigates some failures of DDS in handling multi-domain training.
	\item We demonstrate various exotic use cases of differentiable data selection frameworks (MDAC and original DDS), including domain adaptation using unsupervised clustering instead of using domain label a priori \ref{ssec:clda}, which is barely available in real context; domain adaptation using out-of-domain training data and small in-domain validation data \ref{ssec:uda}; and bi-domain adaptation in which we are only interested in two domain of the training mixture \ref{ssec:bida}.
\end{itemize}

@@

\subsection{Multi-domain NMT}\label{ssec:mda}

We now turn to a more realistic scenario and consider multi-domain NMT, which aims to train one single system with optimal performance averaged over 6 domains. This setting targets a uniform test distribution $\vlambda^t = \vlambda_0$. In this situation, CL \cite{Zhang19curriculum} does not apply. We therefore only contrast the performance of MDAC, DDS and several fixed training data distribution $\vlambda^l \in \big[ \vlambda_0, \vlambda_{0.25}, \vlambda_{0.5}, \vlambda_{0.75}, \vlambda_{1.0}\big]$, where $\vlambda_{\alpha}$ is defined according to equation~\eqref{mixture:trn}.

@@

Multi-domain NMT (MDMT) aims to develop systems that simultaneously bode well for several domains. Like for DA, techniques for supervised MDMT combine one or several ingredients: (a) the specialization of data representations \cite{Kobus17domaincontrol} or of sub-networks \cite{Pham19generic} to differentiate the processing of each domain; (b) the use of adversarial techniques to neutralize differences between domains \cite{Britz17mixing,Zeng18multidomain}; (c) the use of automatic domain identification e.g.\ \newcite{Jiang19multidomain}. Unsupervised MDMT is studied in \newcite{Farajian17multidomain}, as an instance of unsupervised DA.

@@

\newcite{Kumar19reinforcement} use reinforcement techniques (deep Q-learning) to learn the curriculum strategy: in this work, complexity corresponds to difficulty levels which are binned using contrastive data selection. The reward is based on the increase of the development set loss that results from the actual data selection strategy.\fyDone{Alert: what do we do during warm-up ?} The same technique is recently applied to multilingual NMT in \cite{Kumar21learning}. \newcite{Zhou20uncertainty} propose another curriculum-based approach which instead relies on \emph{instance uncertainty} as a measure of their difficulty and presents the data sample starting with the easiest (more predictable) first. Another contribution of this paper is an alternative stopping criterium. Closest to our problems, \newcite{Wang20learning-multi} adapt CL for multi-domain NMT, where an optimal instance weighting scheme is found using Bayesian optimization techniques. Each step consists of (a) weighting instances based on relevance features, (b) fine-tuning a pretrained model using the weighted training set, and is applied iteratively to train a sequence of models. The one that maximizes the development set performance is finally retained.

\section{Conclusion and outlook \label{sec:conclusion}}
\fyDone{Write conclusion}
In this study, we have presented a generic framework to perform a variety of adaptation tasks for machine translation, ranging from conventional supervised domain adaptation to multi-domain NMT and unseen domain adaptation. By experimenting with all these settings, we have shown that the same algorithm, aimed at automatically finding an effective data sampling scheme during the course of training, could be used in all these situations. This algorithm, we believe, provides us with a more sound approach to (multi-domain) DA than existing heuristics and dispenses with the costly search of optimal meta-parameters. Another contribution of our work is an experimental comparison of recent approaches to dynamic data selection.
